---
title: "EDS 214 API Practice"
format: html
editor: visual
---

```{r}
library(dataRetrieval)
library(tidyverse)
```

## USGS Hydrological Data with dataRetrieval

-   Using `dataRetrieval` get the the discharge (ft³/s) time-series for the Ventura River (site `11118500`) during the recent Hilary tropical storm from 2023-08-19 to 2023-08-22

```{r}
# from dataRetrieval CRAN
siteNumber <- "11118500"
parameterCd <- "00060" # Discharge
startDate <- "2023-08-19"
endDate <- "2023-08-22"

discharge <- readNWISdv(siteNumber, parameterCd, startDate, endDate)
```

-   Plot the time-series using `ggplot2`

```{r}
ggplot(data = discharge, aes(x = Date, y = X_00060_00003)) +
  geom_line() +
  geom_point() +
  labs(x = "", 
       y = "Discharge (ft^3 / s)", 
       title = "Ventura River Discharge during Hurricane Hilary")
```

-   What was the maximum discharge during this period of time?

```{r}
max_discharge <- max(discharge$X_00060_00003)

max_discharge
```

## metajam

-   `metajam::download_d1_data`\
    download a specific dataset to your computer using a folder based file structure grouping data and metadata

-   `metajam::read_d1_files`\
    read this data structure to load both data and associated metadata into R using a named list data structure

Let's determine what percentage of Alaskan household are speaking only English!

-   download the data `household_language.csv` using `metajam`

```{r}
# # set inputs
# data_obj <- "https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A9e123f84-ce0d-4094-b898-c9e73680eafa"
# path <- "~/Desktop"
# # download data and metadata
# library(metajam)
# download_d1_data(data_obj, path)
# # Returned
# #[1] "~/Desktop/doi_10.18739_A2DP3X__Alaska_Schools_Rentention2009_15"
```

```{r}
library(metajam)

data_obj <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A7fc6f6db-c5ea-426a-a743-1f2edafb43b8"
path <- "/Users/aherbst/github/EDS214/api-practice" # or just put "." because that means "here"

download_d1_data(data_obj, path)
```

-   Read the data into R using `metajam`

```{r}
# example
# my_data <- read_d1_files("~/Desktop/doi_10.18739_A2DP3X__Alaska_Schools_Rentention2009_15")
```

```{r}
ak_household_data <- read_d1_files("doi_10.5063_F1N58JPP__household_language__csv/")
```

```{r}
household_language <- ak_household_data$data
```

-   compute the percentage of Alaskan household speaking only English from 2009 to 2015

```{r}
percent_english <- household_language %>% 
  filter(Year %in% seq(from = 2009, to = 2015, by = 1)) %>% 
  select(Year, total, speak_only_english, city) %>% 
  group_by(Year) %>% 
  summarize(percent_english_only = sum(speak_only_english) / sum(total),
            total_ppl = sum(total))
```

-   Create a plot to visualize this data

```{r}
ggplot(data = percent_english, aes(x = Year, y = percent_english_only)) +
  geom_col()
```
